---
title: 'Final Project - Step 3'
subtitle: "PSTAT100: Data Science Concepts and Analysis" 

author:
    affiliations:
      - name: "Spring 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```



:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Amelia Hammer(503895-5)
- Katrina Topchiyan(A069255)
- Shreya Reddy(6035174)
- Amelie Machatzke(7357551)
- Meher Brar(3580370)
:::


# Step 3: Final Project Report

## 1.1 Data Analysis
Based on your research questions, dataset characteristics, and hypotheses, choose an appropriate analytical approach. This may involve statistical modeling, hypothesis testing, or in-depth exploratory data analysis using visualizations.

Clearly justify your chosen method(s) and explain how they address your research questions. Where applicable, discuss any assumptions made and their implications.

Possible Approaches:

Regression Analysis: To explore relationships between continuous response and explanatory variables.
Classification: For predicting categorical outcomes (e.g., logistic regression, decision trees).
Clustering: To identify natural groupings within the data.
Time Series Analysis: For analyzing data with temporal structure.
Hypothesis Testing: To statistically test assumptions about your data (e.g., t-tests, chi-square tests, ANOVA).
Exploratory Data Visualization: For uncovering patterns, trends, or anomalies through detailed graphical analysis.


## 1.2 Evaluation and Interpretation

• Use appropriate evaluation techniques based on your selected approach:
• For models, apply relevant performance metrics (e.g., RMSE, accuracy) and include diagnostics to assess validity and assumptions.
• For hypothesis testing, clearly state your null and alternative hypotheses, report p-values or confidence intervals, and interpret the results.
• For visualizations, provide thoughtful interpretation and explain how the visuals support your research findings.

## 1.3 Interpretation of Results

Clearly interpret the results of your models, drawing connections back to your hypotheses and research questions. Discuss any significant patterns, anomalies, or insights found in your analysis.

## 1.4 Visualization and Communication of Findings

Use clear and informative visualizations to support your analysis. Annotate visualizations and discuss their significance. Ensure all visualizations have captions, labels, and are well-documented.

# Final Report

### Abstract
Provide a brief summary of your analysis, key findings, and conclusions in a few sentences.

We are interested in exploring how different car attributes and features affect the way cars are priced. This dataset includes information like brand, model, fuel type, and size, which are important when it comes to a car’s value. These insights could help car sellers make better pricing decisions and assist buyers in understanding what features contribute most to a car’s market value. A potential implication is that we don't know the mileage of these cars from this dataset, which does have an impact on the pricing.

#Still needs Key Findings and conclusions 
#Amelia 

### Introduction
Provide an overview of your dataset, including its source, variables, and the key questions and hypothesis you are investigating.
**About our Data**
This dataset contains information on 205 different car models, including features like car body type, engine size, fuel type, and performance metrics. It includes both categorical and numerical variables, with the target variable being the car's price. The data is useful for analyzing trends in vehicle specifications and building predictive models to estimate car prices.

**Data Name/Title**: Car Price Prediction Multiple Linear Regression
  
  **Author/Owner**: MANISH KUMAR
  
  **Date of Publication**: UPDATED 6 YEARS AGO
  
  **Publication Venue**: Kaggle
  
  **Retrieval Date**: 4/17/25
 
  **Link**: https://www.kaggle.com/datasets/hellbuoy/car-price-prediction

  **Variables**: car_ID, symboling, CarName, fueltype, aspiration, doornumber, carbody, drivewheel, eniginelocation, wheelbase, carlength, carwidth, carheight, curbweight, enginetype, cylindernumber, enginesize, fuelsystem, boreratio, stroke, compressionratio, horsepower, peakrpm, citympg, highwaympg,and price

**Key Questions and Hypothesis**
We are most curious to see how different vehicle characteristics affect price.

We are investigating which vehicle features are the most significant predictors of car price.
Hypothesis 1: Engine size and horsepower are the strongest positive predictors of price.

We are also exploring how a car’s physical dimensions, specifically its length, affect its price.
Hypothesis 2: Longer vehicles, on average, have higher prices compared to shorter vehicles.

We are examining whether there are statistically significant differences in car prices across body types (e.g., sedan, hatchback, convertible). 
Hypothesis 3: Convertibles have the highest average prices, while hatchbacks have the lowest average prices.

### Data Processing
Describe the steps taken for data cleaning, handling missing values, and any necessary preprocessing. Mention any transformations or feature engineering applied to improve the model’s performance.
#Amelia 

### Modeling Process and Results
Describe your approach to selecting and implementing models.
Present your key results, using visualizations and tables as needed.
#Katrina, Shreya, and Amelie

**HYPOTHESIS 1**
Engine size and horsepower are the strongest positive predictors of price.

To evaluate the hypothesis that engine size and horsepower are the strongest positive predictors of car price, we selected a multiple linear regression model. This model is appropriate because the response variable, price, is continuous, and the explanatory variables—engine size and horsepower—are also continuous and expected to have a linear relationship with price. Multiple linear regression allows us to estimate the individual contribution of each predictor while controlling for the other. This is essential in our case, since engine size and horsepower may be correlated, and we want to isolate their independent effects on price.

```{r}
# model for hypothesis 1: multiple linear regression predicting price from engine size + horsepower
model1 <- lm(price ~ enginesize + horsepower, data = df)
summary(model1)
```
-all p-vales are <.001, which means both predictors are highly statistically significant
-engine size and horse power are both strong, signifcant, positive predictors of price 
-R-sqaured: .7933 so about 79.3% of the variation in care price is explained by this model 
-Engine size has a coefficient of 122.45 and horsepower a coefficient of 58.85. This indicates that, on average, every one-unit increase in engine size or horsepower corresponds to an increase in car price by $122.45 and $58.85, respectively, holding the other variable constant

```{r}
cor(df$enginesize, df$price)
cor(df$horsepower, df$price)
```
both correlations are reasonably high

```{r}
install.packages("caret")
library(caret)
varImp_model <- train(price ~ enginesize + horsepower + carlength + carwidth + carheight + curbweight + citympg + highwaympg + peakrpm, data = df, method = "lm")
plot(varImp(varImp_model))
```
```{r}
model_full <- lm(price ~ enginesize + horsepower + carlength + carwidth + carheight + curbweight + citympg + highwaympg + peakrpm, data = df)
summary(model_full)
```
Based on the visualization and the regression summary, we can see that horsepower is not one of the highest predictors of price when controlling for other variables. Although it is statistically significant (p ≈ 0.038), its relative influence is smaller compared to variables like engine size, curbweight, and carwidth. This suggests that while horsepower contributes to price prediction, it is not among the most dominant factors in determining a car's value in the presence of other influential features.

**HYPOTHESIS 2**
Longer vehicles, on average, have higher prices compared to shorter vehicles.

Two-Sample t-test

We used a two-sample t-test because our hypothesis compared two groups, long and short cars. The t-test allowed us to directly compare the means of each group and conclude whether the mean price of each group was statistically significant.

```{r}
df$is_long <- ifelse(df$carlength > 196, "Long", "Short")
t.test(price ~ is_long, data = df)
```
The two-sample t-test demonstrated a statistically significant difference in average car prices between long and short vehicles. Cars with length greater than 196 inches were considered long, and cars with length shorter than 196 inches were considered short. The t-test returned a p-value of 0.0008507, which is less than $\alpha$ = 0.05, meaning there is no strong evidence to reject the null hypothesis that there is a statistically significant difference between the prices of long and short cars. Long cars have a mean price of $28,808.55 and short cars have a mean price of $12396.04, indicating that longer cars tend to be significantly more expensive.

Regression

We used a simple linear regression in order to assess the continuous relationship between car length and price and measure the slope of the best fit line. This supplemented our findings from the t-test, since the t-test only compared cars greater than and less than 196 inches long.

```{r}
model_length <- lm(price ~ carlength, data = df)
summary(model_length)
```
Since the p-value is < 2.2e-16, the simple linear regression model shows a statistically significant, positive relationship between car length and price. Each additional inch in car length is associated to an average increase of $442 in car price.

```{r}
plot(df$carlength, df$price,
     main = "Car Length vs. Price",
     xlab = "Car Length (inches)",
     ylab = "Price (USD)",
     pch = 16,
     col = "royalblue4")

abline(model_length, col = "red", lwd = 2, lty = 2)
```
The scatter plot of Car Length vs. Price shows a clear positive trend, demonstrating that greater car length is positively correlated with car price.

**HYPOTHESIS 3**
Convertibles have the highest average prices, while hatchbacks have the lowest average prices.

To address the research question — Are there statistically significant differences in car prices across body types? — we chose one-way ANOVA, a statistical hypothesis testing method suitable for comparing means of a continuous outcome (price) across categories (carbody types). This model fits our goal of evaluating whether certain car body types (like convertibles and hatchbacks) tend to have significantly different price levels. One-way ANOVA tests the null hypothesis that the mean car price is the same for all body types. If significant, we further apply Tukey’s HSD post-hoc test to determine which specific groups differ.
# Visualizing Mean Prices
```{r}
library(ggplot2)
ggplot(df, aes(x = carbody, y = price)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Car Price Distribution by Body Type", y = "Price", x = "Car Body Type") +
  theme_minimal()
```
The boxplot shows clear differences in median price across car body types. Convertibles and hardtops have the highest price potential, while hatchbacks have the most compressed and lowest price distribution. This gives preliminary visual support for Hypothesis 3.

# Running One-Way ANOVA
```{r}
anova_model <- aov(price ~ carbody, data = df)
summary(anova_model)
```
The p-value for `carbody` is 0.0000145, which is highly significant. This means we reject the null hypothesis that all body types have equal mean prices. There are statistically significant differences in car prices across body types.

# Finding Which Specific Groups Differ Using Post-Hoc Test (Tukey’s HSD)
```{r}
TukeyHSD(anova_model)
```
The Tukey’s HSD test identifies which specific pairs of car body types differ significantly. The comparison between convertibles and hatchbacks is statistically significant, confirming that convertibles are priced higher than hatchbacks. To further confirm Hypothesis 3 that convertibles have the highest average prices while hatchbacks have the lowest average prices, we look at the mean prices summary.

# Mean Prices Summary
```{r}
aggregate(price ~ carbody, data = df, mean)
```
This table confirms that convertibles have the highest average price and hatchbacks have the lowest average price, proving that Hypothesis 3 is true. Convertibles had a price of 20391.15 dollars and hatchbacks had the price of 10391.97 dollars. These numerical findings match what was observed visually and statistically.


### Interpretation

**HYPOTHESIS 1**
Engine size and horsepower are the strongest positive predictors of price.
Both engine size and horsepower are significant predictors of car price which we can see in the initial multiple linear regression model including only these two variables, both were found to have strong positive coefficients and p-values well below 0.001.

However, when we added other variables to the model, such as car length, width, height, curb weight, fuel efficiency metrics, and peak RPM, we observed that the relative impact of horsepower decreased. Although it remained statistically significant, its effect size was smaller than that of other features like curb weight and car width. 

These findings demonstrate that engine size is a consistently strong and influential predictor, while horsepower is important but comparatively less impactful in the presence of other variables. This partially confirms our hypothesis and highlights the value of using a multivariate approach to isolate the unique contribution of each predictor.

**HYPOTHESIS 2**
Longer vehicles, on average, have higher prices compared to shorter vehicles.

We used both a two-sample t-test and simple linear regression. The t-test showed a statistically significant difference in average prices between long cars (longer than 196 inches) and short cars (shorter than 196 inches), with p = 0.00085. On average, longer vehicles cost approximately $16,412.51 more than short vehicles, with the average short vehicle price being $12396.04, and average long vehicle price being $28,808.55.

The simple regression analysis confirmed a strong positive relationship between car length and price, showing that each additional inch in length is associated with about a $442 increase in price. 

Therefore, car length is positively correlated with price.

**HYPOTHESIS 3**
Convertibles have the highest average prices, while hatchbacks have the lowest average prices.

Our analysis strongly supports Hypothesis 3. Using one-way ANOVA, we found statistically significant differences in mean car prices across body types. The Tukey post-hoc test further confirmed that convertibles are significantly more expensive than hatchbacks. These results answer the research question with statistical backing and suggest that body type is an important driver of car pricing.

These findings highlight how vehicle body type can serve as a meaningful factor when analyzing or predicting car price variability. While the model does not capture all variation in price, it demonstrates that carbody is an important categorical predictor worth exploring further in more complex models.

### Visualization and Communication
Include well-labeled and informative visualizations, with captions explaining their relevance. Ensure clarity and professionalism in all graphical elements.
#Meher

### Conclusion and Recommendations
Summarize your key insights and provide actionable recommendations.Discuss any limitations of your analysis and suggest potential future directions.
#Amelia and Meher 



